Abstractions for Multi-Sorted Substitutions

### 2023-09-18 paper/paper.tex
Update.

  The rest of this paper is structured as followed:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to introduce the notion of types
  for multi-sorted syntax and how to define typing contexts.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservattion lemmas for renamings and substitutions.
  Section~\ref{sec:semantics} introduces a small-step operational semantics for System F.
  Section~\ref{sec:subjectreduction} proves the subject reduction lemma for System F.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:casestudies} describes case studies of object languages beyond System F.
  Section~\ref{sec:future} describes future work.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:conclusion} concludes.

@@

  \section{Future Work}
  \label{sec:future}
  \begin{itemize}
  \item
    Enabling more object languages by allowing subterms with less
    restrictions on the form of \ACode{S}.
  \item
    Providing abstractions not just for type-preservation, but also
    for different kinds of semantics, e.g. confluence or lemmas related
    to denotational semantics.
  \end{itemize}

@@

  The kit abstraction for unifying renamings and substitutions
  appeared first in an unpublished article by
  McBride\cite{unpublished:mcbride2005kits}, and then later in
  Benton et al.\cite{DBLP:journals/jar/BentonHKM12}. 
  Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} propose an
  extension to kits that supports substructural types via resource
  vectors.
  In all three cases they are formulated for intrinsic typing.

  \emph{Autosubst}\cite{DBLP:conf/itp/SchaferTS15}
  is a Coq framework, which derives parallel substitution definitions and
  lemmas for languages from annotated Coq syntax definitions using
  extrinsic typing, extrinsic scoping, and de Bruijn indices.
  The framework is implemented in Coq's tactic language LTAC and comes
  with a decision procedure for all assumption-free, equational
  substitution-lemmas.
  The implementation of Autosubst deals with multiple variable sorts
  by generating multiple substitutions and corresponding interaction lemmas.
  Autosubst does not provide any machinery for type preservation and
  does not unify renaming and substitution.
  % Assumes fun-ext.

  \emph{Autosubst 2}\cite{DBLP:conf/cpp/StarkSK19} is a standalone
  code generator, which translates second-order HOAS specifications into
  mutual inductive term sorts. Compared to Autosubst 1, it features
  mutually recursive object languages, intrinsically scoping, and
  vectorized substitutions.
  They also define and prove lemmas for all four compositions.
  The generated syntax consists of multiple term types, indexed by
  different amounts of free variables as shown in Section~\ref{sec:syntax}.
  % Does not assume fun-ext.

  Allais et al.\cite{DBLP:conf/cpp/Allais0MM17} propose a powerful
  abstraction for denotational semantics and proving semantic fusion
  lemmas.
  In later work\cite{DBLP:journals/pacmpl/AllaisA0MM18}, they use
  generic programming to instanciate this abstraction for a class of
  object languages comparable to ours.
  They demonstrate how both renaming and substitutions can be
  described as semantics and how the four composition laws follow from
  their generic fusion laws.
  They do not show how to work with multiple variable sorts, how to
  abstract over the compositions, and how to abstract over typing in
  an extrinsically typed scenario.

  Autosubst \cite{DBLP:conf/cpp/StarkSK19, DBLP:conf/itp/SchaferTS15}.
  Kits \cite{DBLP:journals/jar/BentonHKM12, unpublished:mcbride2005kits}.
  Kits linear \cite{DBLP:journals/corr/abs-2005-02247}.
  Universe \cite{DBLP:journals/pacmpl/AllaisA0MM18}.

@@

>>>>

  The rest of this paper is structured as followed:
  Section~\ref{sec:syntax} introduces the multi-sorted syntax and compares it to the
  more common unsorted syntax.
  Section~\ref{sec:maps} introduces multi-sorted substitutions and renamings, and
  an abstraction to unify them.
  Section~\ref{sec:composition} introduces composition of multi-sorted
  substitions and renamings, and an abstraction to unify all four
  compositions.
  Section~\ref{sec:types} shows how to introduce the notion of types
  for multi-sorted syntax and how to define typing contexts.
  Section~\ref{sec:typing} shows how to define a multi-sorted typing relation,
  and an abstraction to unify type preservattion lemmas for renamings and substitutions.
  Section~\ref{sec:semantics} introduces a small-step operational semantics for System F.
  Section~\ref{sec:subjectreduction} proves the subject reduction lemma for System F.
  Section~\ref{sec:reflection} describes the class of object languages covered by
  our reflection algorithm.
  Section~\ref{sec:casestudies} describes case studies of object languages beyond System F.
  Section~\ref{sec:related} discusses related work.
  Section~\ref{sec:future} describes future work.
  Section~\ref{sec:conclusion} concludes.

@@

@@

  The kit abstraction for unifying renamings and substitutions
  appeared first in an unpublished manuscript by
  McBride\cite{unpublished:mcbride2005kits}, and then later in
  Benton et al.\cite{DBLP:journals/jar/BentonHKM12}. 
  Wood and Atkey\cite{DBLP:journals/corr/abs-2005-02247} propose an
  extension to kits that supports substructural types via resource
  vectors.
  In all three cases they are formulated for intrinsic typing.

  \emph{Autosubst}\cite{DBLP:conf/itp/SchaferTS15}
  is a Coq framework, which derives parallel substitution definitions and
  lemmas for languages from annotated Coq syntax definitions using
  extrinsic typing, extrinsic scoping, and de Bruijn indices.
  The framework is implemented in Coq's tactic language LTAC and comes
  with a decision procedure for all assumption-free, equational
  substitution-lemmas.
  The implementation of Autosubst deals with multiple variable sorts
  by generating multiple substitutions and corresponding interaction lemmas.
  Autosubst does not provide any machinery for type preservation and
  does not unify renaming and substitution.

  \emph{Autosubst 2}\cite{DBLP:conf/cpp/StarkSK19} is a standalone
  code generator, which translates second-order HOAS specifications into
  mutual inductive term sorts. Compared to Autosubst 1, it features
  mutually recursive object languages, intrinsically scoping, and
  vectorized substitutions.
  They also define and prove lemmas for all four compositions.
  The generated syntax consists of multiple term types, indexed by
  different amounts of free variables as shown in Section~\ref{sec:syntax}.

  Allais et al.\cite{DBLP:conf/cpp/Allais0MM17} propose a powerful
  abstraction for denotational semantics and semantic fusion lemmas.
  In later work\cite{DBLP:journals/pacmpl/AllaisA0MM18}, they use
  generic programming to instanciate this abstraction for a class of
  object languages comparable to ours.
  They demonstrate how both renamings and substitutions can be
  described as semantics and how the four composition lemmas follow from
  their generic fusion lemma.
  They do cover an abstraction to unify renamings and substitutions.
  They do not show how to work with multiple variable sorts, how to
  abstract over the compositions, and how to abstract over typing in
  an extrinsically typed scenario.
  However, we believe that their framework is general enough such that
  our concepts can also be transferred to their framework.

@@

  \section{Future Work}
  \label{sec:future}
  \begin{itemize}
  \item
    Enabling more object languages by allowing subterms with less
    restrictions on the form of \ACode{S}, allowing recursive
    occurences in container types like lists and finite functions, and
    adding support for substructural types as in
    \cite{DBLP:journals/corr/abs-2005-02247}.
  \item
    Incorporating a decision procedure for equational
    substitution-lemmas based on $\sigma$-calculus as in Autosubst
    \cite{DBLP:conf/itp/SchaferTS15, DBLP:conf/cpp/StarkSK19}.
  \item
    Developing abstractions and derived lemmas beyond substition and
    type preservation, e.g. for confluence or denotational semantics
    as in \cite{DBLP:journals/pacmpl/AllaisA0MM18}.
  \end{itemize}
